---
title: "Introduction to Geospatial Techniques for Social Scientists in R"
subtitle: "Vector Data"
author: "Stefan Jünger & Anne-Kathrin Stroppe"
date: June 06, 2022
institute: |
  GESIS Workshop
presenter: Anne
editor_options: 
  chunk_output_type: console
---
layout: true

```{r child = "./content/config/sessions_setup.Rmd"}
```

---

## Now

```{r course-content-now, echo = FALSE}
course_content %>%
  kableExtra::row_spec(2, background = "yellow")
```

---

## Why care about data types and formats?

There is a difference in how spatial information is stored, processed, and visually represented.
- Different commands to import and handle the data
- Spatial linking techniques and analyses partially determined by the data format
- Visualization of data can differ

So: Always know what kind of data you are dealing with!

---

## Representing the world in vectors

.pull-left[
.center[
```{r world-cities, echo = F, out.width= "120%"}
data(World, metro)

tm_shape(World) +
  tm_borders()+
  tm_shape(metro) +
  tm_dots()
```
]
]

.pull-right[
The surface of the earth is represented by simple geometries and attributes.

What usually comes as vector data:
- administrative borders 
- rivers and roads, 
- buildings, cities, and more

Every object is defined by longitude (x) and latitude (y) values.
]

---

## Vector data: Geometries
.pull-left[
Each real-world feature is one of three types of geometries:
- Points: discrete location (e.g., city)
- Lines: linear feature (e.g., river)
- Polygons: enclosed area  (e.g, country)
]

.pull-right[
```{r vector-geo, echo = FALSE, out.width = "90%"}
woRkshoptools::include_picture("vector_geometries.png")
```
<br>
<small><small><small> National Ecological Observatory Network (NEON), cited by [Datacarpentry](https://datacarpentry.org/organization-geospatial/02-intro-vector-data/)</small></small></small>
]

---

## Vector data: Attribute tables

The geometries of vector data hold up to three information:
- location (points, lines, and polygons)
- length (lines and polygons)
- area (polygons)

We need to assign attributes to each geometry to hold additional information $\rightarrow$ data tables called attribute tables
- Each row represents a geometric object, which we can also call observation or case.
- Each column holds an attribute or, in "our" language, a variable.

---

## Vector data: Attribute tables

.center[
```{r attr-table, echo = FALSE, out.width = "90%"}
woRkshoptools::include_picture("attr_table.png")
```
]

---

## New best friend: Shapefiles

Both the geometric information and attribute table can be saved within one file. Rather often, *ESRI Shapefiles* are used to store vector data.
Shapefiles consist of at least three mandatory files with the extensions:
- .shp : shape format
- .shx : shape index format
- .dbf : attribute format
- (.prj: CRS/Projection)

You don't have to remember what they are standing for, but know that you won't be able to load the data if one of those files is missing.


---

## Welcome to `simple features`

.pull-left[
.small[
Several packages are out there to wrangle and visualize spatial and, especially, vector data within `R`. We will use a package called `sf` ("simple features").

Why? 
`simple features` refers to a formal standard representing spatial geometries and supports interfaces to other programming languages and GIS systems.
]
]

.pull-right[
```{r echo = FALSE}
woRkshoptools::include_picture("sf.jpg")
```
<small><small>Illustration by [Allison Horst](https://github.com/allisonhorst/stats-illustrations) </small></small>
]

---

## Load a shapefile

The first step is, of course, loading the data.
We want to import the shapefile the administrative borders of the German states (*Bundesländer*) called `VG250_LAN.shp`.


```{r load-germanstates-display}
# load library
library(sf)

german_states <- sf::read_sf("./data/VG250_LAN.shp")
```

---

## Inspect your data: Classics

Let's have a quick look at the imported data.
Like every other data set, we inspect the data to check some metadata and see if the importing worked correctly.

```{r data-inspection-classy}
# object type
class(german_states) 

# number of rows
nrow(german_states)

# number of columns
ncol(german_states)
```

---

## Inspect your data: Classics

You can see that there are no huge differences between the shapefile, we just imported, and a regular data table.

```{r data-inspection-head}
# head of data table
head(german_states, 2)
```

---

## Inspect your data: Spatial features

Besides our general data inspection, we also want to check the spatial features of our import.
This includes the geometric type (points? lines? polygons?) and the coordinate reference system.

```{r data-inspection-stgeo}
# type of geometry
sf::st_geometry(german_states) 
```


---
## Inspect your data: Spatial features

Each polygon is defined by several points that are connected to build an enclosed area. Several polygons in one data frame have the `sf` type multipolygons. Just as Germany consists of several states, the polygon Germany consists of several smaller polygons.

```{r data-inspection-stattr}
# the simple features column
attr(german_states, "sf_column") 

# further inspecting 
dplyr::glimpse(german_states$geometry)
```

---

## Inspect your data: Spatial features

Remember: The Coordinate Reference System is very important. A crucial step is to check the CRS of your geospatial data.

```{r data-inspection-crs}
# coordinate reference system
sf::st_crs(german_states) 
```

---

## `sf::st_transform()`

When a CRS is messed up, or one wants to combine data with non-matching CRS, it will all go downwards.
The good thing is that the command `sf::st_transform()` allows us to  *translate* our spatial data from one coordinate reference system to another.

```{r st-transform}
# transform crs
german_states <- sf::st_transform(german_states, crs = 3035)

# check crs
sf::st_crs(german_states) 
```

---

## A very, VERY first map

For inspecting the data and check if we actually loaded what we want to load, we can have a very first glimpse.

```{r first-glimpse, eval = F}
# plot sf object
plot(german_states) 
```

.center[
```{r plot-states, echo = FALSE, out.width = "60%"}
woRkshoptools::include_picture("plot_german_states.png")
```
]

---

## Import point layer

Unfortunately, the data we want to visualize or analyze are not always available as shapefiles.
Especially, point coordinates are often stored in table formats like `.csv` - as are the location of German hospitals in our `./data` folder. 

```{r load-hospitals-display}
hospitals_df <- readr::read_csv("./data/hospital_points.csv")
head(hospitals_df)
```


---

## From data table to geospatial data

We see that besides our attributes (e.g., year of data collection, number of beds...), the table contains the two variables "X" and "Y", our point coordinates.
When using the command `sf::st_as_sf()`, make sure to use the option `crs = [EPSG_ID]`.
If not used, your CRS will not be defined, and you won't perform further commands depending on the CRS.
Good thing that Anne was the one who geocoded the addresses, and we know that the data are most likely stored in `EPSG:3035`.

.pull-left[
```{r plot-hospitals-sf-display, eval = F}
# transform to spatial data frame
hospitals_sf <- 
  sf::st_as_sf(
    hospitals_df,    
    coords = c("X", "Y"),
    crs = 3035
  )

# inspect data
class(hospitals_sf)
sf::st_geometry(hospitals_sf)
```
]

.pull-right[
```{r transform-hospitals, echo = F}

# transform to spatial data frame
hospitals_sf <- 
  sf::st_as_sf(
    hospitals_df,    
    coords = c("X", "Y"),
    crs = 3035
  )
# inspect data
class(hospitals_sf)
sf::st_geometry(hospitals_sf) 
```
]

---

## Our point data

```{r plot-hospitals, out.width = "55%"}
plot(hospitals_sf)
```

---

## ... and the other way round

Do you want to go back to handling a simple data frame? You can quickly achieve this by drop the geometry column.

```{r remove-geometry}
# check class
class(german_states)

# remove geometry
sf::st_drop_geometry(german_states) %>% 
  head(2)
```

---

class: middle
## Exercise 1_2_1: Import Vector Data 

[Exercise](https://stefanjuenger.github.io/gis_socsci_konstanz/exercises/1_2_1_Import_Vector_Data.html)

[Solution](https://stefanjuenger.github.io/gis_socsci_konstanz/solutions/1_2_1_Import_Vector_Data.html)

---

## Data wrangling

After importing the data sets, we are now ready to manipulate our data. 
We are working here with the `dplyr` package to manipulate the data frames for all regular data wrangling tasks.
But if you are used to working with the base R language, feel free to do so.

.center[
```{r echo = FALSE, out.width = "50%" }
woRkshoptools::include_picture("tidyverse.png")
```
<small><small>Meme found on [Reddit](https://www.reddit.com/r/Rlanguage/comments/anv1d5/my_meme_of_the_day/?utm_source=share&utm_medium=web2x&context=3) </small></small>
]

---

## Data Intro: German Districts

We're moving "a layer down" and look at Germany on a more fine-grained spatial level: the district.

```{r load-district-display}
german_districts <- 
  sf::read_sf("./data/VG250_KRS.shp") %>% 
  sf::st_transform(crs = 3035) 

german_districts
```


---
## German Districts

.small[
```{r plot_districts, echo = F , out.width = "70%"}
plot(german_districts)
```
]
---

## Data Intro: Attributes

Since it would be a little bit boring to work with just administrative information, there's an extra table with more attributes called *attributes_districts.csv*.

```{r load-attr-display}
attributes_districts <- readr::read_csv("./data/attributes_districts.csv") 
```



```{r head-district-attr, echo = F, out.width= "60%"}
head(attributes_districts, 2)
```

---

## Add attributes: Join data table

You might already spot that we have an id for the districts (*AGS* and *district_id*) in both data tables. 
This is a good example how easily `sf` objects can be handled.

```{r join-attrtable}
german_districts_enhanced <- 
  german_districts %>% 
  dplyr::rename(district_id = AGS) %>% 
  dplyr::select(district_id) %>% 
  dplyr::left_join(attributes_districts, by = "district_id")  

head(german_districts_enhanced, 2)
```

---

## Add (More) Attributes: Spatial Join, Spatial Intersects, and More

Besides the regular join, we can also perform a so-called *spatial join*.
For example, we want to count the number of hospitals in each German district.

```{r spatial-join}

# perform spatial join to identify for each hospital the correct district id 
hospitals_in_districts <- 
  sf::st_join(hospitals_sf, german_districts_enhanced, join = st_within)

# count the number of hospitals within a district
hospital_districts_count <- 
  dplyr::count(hospitals_in_districts, district_id, name = "hospital_count")

# Join the hospital count with the German district a
german_districts_enhanced <- 
  dplyr::left_join(
    german_districts_enhanced, 
    hospital_districts_count %>% sf::st_drop_geometry(), 
    by = "district_id"
  ) %>% 
  dplyr::mutate(hospital_count = tidyr::replace_na(hospital_count, 0))
```

---

## Subsetting the Data

One might be interested in only one specific area of Germany, like the capital Berlin.
To subset a `sf` object,  you often can use your usual data wrangling workflow. 
In this case, I know the district_id and that is also the only information I want to keep.

.pull-left[
```{r stouches-display, eval = F}
berlin <-
  german_districts_enhanced %>% 
  dplyr::filter(district_id == 11000) %>% 
  dplyr::select(district_id) 

plot(berlin)
```
]

.pull-right[
```{r stouches, echo = F}
# subsetting
berlin <-
  german_districts_enhanced %>% 
  filter(district_id == 11000) %>% 
  dplyr::select(district_id) 

plot(berlin)
```
]
---

## Using `sf` for subsetting

If you have no information about *ids* but only about the geolocation, you can use `sf::st_touches()` (or `st_touches()`, `st_within()`, `st_intersect()`, `st_crosses()`...) to identify for example all districts which share a border with Berlin.

.pull-left[
```{r surround-display, eval = F}
berlin_surrounding <-
  german_districts_enhanced %>% 
  dplyr::select(district_id) %>% 
  # length of mutual border > 0
  dplyr::filter(
    lengths(sf::st_touches(., berlin)) > 0
  ) 

plot(berlin_surrounding)
```
]

--

.pull-right[
```{r surround, echo= F}
berlin_surrounding <-
  german_districts_enhanced %>% 
  dplyr::select(district_id) %>% 
  # length of mutual border > 0
  dplyr::filter(
    lengths(sf::st_touches(., berlin)) > 0
  ) 

plot(berlin_surrounding)
```
]

---

## Export the Data

After Wrangling and adjusting the data, you can save them.
There are, again, several options to do so.
Two notes:
.small[

1.Be careful when saving shapefiles: column names will automatically be abbreviated!

2.Make sure that the CRS is included either in your folder or the file name.

]


```{r export-display, eval = F}

# Export as Shapefile
sf::st_write(
  german_districts_enhanced, 
  "./participant_materials/districts_enhanced_epsg3035.shp"
)

# Export data frame as csv without geometric attributes
german_districts_enhanced_df <- 
  sf::st_drop_geometry(german_districts_enhanced) 

readr::write_csv(
  german_districts_enhanced, 
  "./participant_materials/german_districts_enhanced.csv"
)

# Export data frame as csv with geometric attributes
sf::st_write(
  hospitals_sf, "./participant_materials/hospitals_epsg3035.csv", 
  layer_options = "GEOMETRY=AS_XY"
)
```

---

class: middle
## Exercise 1_2_2: Manipulate Vector Data 

[Exercise](https://stefanjuenger.github.io/gis_socsci_konstanz/exercises/1_2_2_Manipulate_Vector_Data.html)

[Solution](https://stefanjuenger.github.io/gis_socsci_konstanz/solutions/1_2_2_Manipulate_Vector_Data.html)

---

## Wrap-Up

.pull-left[
We made it through our first session dealing with vector data!
You can:
- load
- transform
- manipulate 
- and export vector data.

The next step is producing an awesome map! `r emo::ji("confetti")`
]

.pull-right[
```{r cologne-covid, echo=F}

# Import data
german_districts <- 
  sf::read_sf("./data/VG250_KRS.shp") %>% 
  dplyr::rename(district_id = AGS)

attributes_districts <- readr::read_csv("./data/attributes_districts.csv") 

# Join data and transform
german_districts_enhanced <- 
  german_districts %>% 
  dplyr::left_join(attributes_districts, by = "district_id") %>% 
  sf::st_transform(3035)

# filter Mannheim
mannheim <-
  german_districts_enhanced %>% 
  dplyr::filter(district_id == "08222")

# filter surrounding districts, append with Mannheim data and select the Covid column
mannheim_sur <-
  german_districts_enhanced %>%
  dplyr::filter(lengths(sf::st_touches(., mannheim)) > 0) %>% 
  dplyr::bind_rows(mannheim) %>%   
  dplyr::select(covid_deaths_pop)

# plot  
plot(mannheim_sur)
```
]

---

class: middle
## Lunch Break `r emo::ji("taco")`
.center[
But, that is for after lunch!
]
---

```{r child = "./content/config/sessions_end.Rmd"}
```