---
title: "Introduction to Geospatial Techniques for Social Scientists in R"
subtitle: "Advanced Data Import & Processing"
author: "Stefan Jünger & Anne-Kathrin Stroppe"
date: June 07, 2022
institute: |
  GESIS Workshop
presenter: Stefan
editor_options: 
  chunk_output_type: console
---
layout: true

```{r child = "./content/config/sessions_setup.Rmd"}
```

---

## Now

```{r course-content-now, echo = FALSE}
course_content %>%
  kableExtra::row_spec(6, background = "yellow")
```

---

## `R`-Packages *We* Will Use

- `osmdata` 
- `tmaptools`
- `StefanJuenger/z11`
- (`reticulate`)

---

## More on the Geospatial Data Landscape

Geospatial data tend to be quite big
- pressure to distribute data efficiently

Data dumps (on the internet) may not be helpful when resources are low and/or time's a factor.

Already for a long time, there are standards to distribute data over the internet and make accessing only chunks of them possible
- Use of *Programming Application Interfaces* (API)

---

## What Is an API?

An Application Programming Interface (API) serves as an entry point to data distributed over the internet to
- get data
- push data

Standardized mechanisms
- to query data
- often just a simple text string to enter in your URL-address bar
- it gets complicated when login data are required

---

## Data Providers Offering Geospatial Data APIs

- [OpenStreetMap](https://wiki.openstreetmap.org/wiki/API)
- [Google](https://developers.google.com/maps/documentation/geolocation/overview)
- [Bing](https://docs.microsoft.com/en-us/bingmaps/rest-services/locations/)
- ...
- [Cologne's Open Data Portal](https://www.offenedaten-koeln.de/dataset/taxonomy/term/44/field_tags/Geo-44)
- Specialized `R` packages, such as the [`wiesbaden` package](https://cran.r-project.org/web/packages/wiesbaden/index.html) or the [`tidycensus` package](https://cran.r-project.org/web/packages/tidycensus/index.html)

---

## How Are Data Accessed?

1. (What is there to get?)
2. (Can I get it?)
3. Specify what I can get
3. Get it!

---

## Example: Access to Public Transport

Say, we're interested in the accessibility of public transport in Cologne
- bus, tram, etc.
- all platforms and vehicles should be wheel-chair accessible

**We can gather this information using OpenStreetMap!**

---

## Accessing OSM Data: The Overpass API

> The Overpass API (formerly known as OSM Server Side Scripting, or OSM3S before 2011) is a read-only API that serves up custom selected parts of the OSM map data. It acts as a database over the web: the client sends a query to the API and returns the data set that corresponds to the query.

.tinyisher[Source: https://wiki.openstreetmap.org/wiki/Overpass_API]

---

## Starting With a Geographic Area to Query

Many geospatial API requests start with a bounding box or another geographical extent to define which area should be accessed.

.pull-left[
```{r cologne-polygon}
cologne_pt_stops <-
  osmdata::getbb(
    "Köln"
  )
```
]

--

.pull-right[
```{r cologne-polygon-plot, echo = FALSE} 
cologne_pt_stops %>% 
  t() %>% 
  dplyr::as_tibble() %>% 
  sf::st_as_sf(coords = c("x", "y")) %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() %>% 
  tm_shape() +
  tm_polygons()
```
]

---

## Defining Some Technical Details

The Overpass API also requires a timeout parameter that repeats the request a certain amount of time if anything fails.

```{r cologne-opq}
cologne_pt_stops <-
  cologne_pt_stops %>% 
  osmdata::opq(timeout = 25*100)

cologne_pt_stops
```

---

## Turning to the Content

The content which we aim to request is defined with key/value pairs. It's best to learn them by doing and look them up in the [official documentation](https://wiki.openstreetmap.org/wiki/Map_features).

```{r cologne-features}
cologne_pt_stops <-
  cologne_pt_stops %>%    
  osmdata::add_osm_feature(key = "public_transport", value = "stop_position")

cologne_pt_stops
```

---

## Conduct Actual Request/Download

In the `osmdata` package, data is actually downloaded, e.g., by using the `osmdata::osmdata_sf()` function.

```{r osmdata-download}
cologne_pt_stops <-
  cologne_pt_stops %>%   
  osmdata::osmdata_sf()

cologne_pt_stops
```

---

## Filter and Transform

The data comprises a list that can be accessed as any list in `R`. Here we extract the points and wrangle them spatially.

.pull-left[
```{r osmdata-filter-transform}
cologne_pt_stops <-
  cologne_pt_stops %>% 
  .$osm_points %>% 
  tibble::as_tibble() %>% 
  sf::st_as_sf() %>% 
  sf::st_transform(3035) %>% 
  dplyr::filter(wheelchair == "yes")

cologne_pt_stops
```
]

---

## The Data Indeed Are Mappable

.pull-left[
```{r cologne-pt-stops-map, eval = FALSE}
tm_shape(cologne_pt_stops) +
  tm_dots()
```
]

--

.pull-right[
```{r cologne-pt-stops-map, ref.label = "cologne-pt-stops-map", echo = FALSE}
```
]

---

## Fiddling With the Data: Creating a Quick 'Heatmap'

OpenStreetMap points data are nice to analyze urban infrastructure. Let's draw a quick 'heatmap' using kernel densities.

.pull-left[
```{r cologne-pt-stops-densities}
cologne_pt_stops_densities <- 
  cologne_pt_stops %>% 
  sf::as_Spatial() %>%
  as("ppp") %>% 
  spatstat.explore::density.ppp(sigma = 500) %>% 
  terra::rast()

terra::crs(cologne_pt_stops_densities) <- "epsg:3035"

```
]

--

.pull-right[
```{r cologne-pt-stops-densities-plot, echo = FALSE}
tm_shape(cologne_pt_stops_densities) +
  tm_raster(midpoint = NA, palette = "-viridis")
```
]

---

class: middle
## Exercise 2_1_1: Working with OSM Data

[Exercise](https://stefanjuenger.github.io/gis_socsci_konstanz/exercises/2_1_1_Working_with_OSM_data.html)

[Solution](https://stefanjuenger.github.io/gis_socsci_konstanz/solutions/2_1_1_Working_with_OSM_data.html)

---

## Accessing Unpackaged (Vector) Data

Not all data come as pretty as OpenStreetMap data in, e.g., the `osmdata` package.

Don't worry. There are methods to import data from source that
- only provide an URL
- not yet prepared for analysis


---

## Example: GeoJSON Files

JSON files are pretty popular
- standardized and well structured
- similar to XML-files, but human readability is a bit better
- also easy to parse for editors and browser

There's also a JSON flavor for geospatial data...

---

## GeoJSON Snippet

```
{
  "type": "FeatureCollection",
  "features": [
    {
      "type": "Feature",
      "id": 12,
      "geometry": {
        "type": "Polygon",
        "coordinates": [
          [
            [
              6.957362270020273,
              50.94308762750329
            ]   
  ...
```

.tinyisher[Source: https://www.offenedaten-koeln.de/]

---

## An Application From Cologne’s "Sag's uns" Portal

```{r sags-uns-cologne}
sags_uns_cologne <-
  glue::glue(
    "https://sags-uns.stadt-koeln.de/sites/default/files/geojson/reports.json"
  ) %>% 
  sf::st_read(as_tibble = TRUE)
```

.tinyisher[Source: https://sags-uns.stadt-koeln.de/]

---

## Trash Complaints

With just two 'simple' command we can retrieve geospatial data about Cologne's inhabitants complaining about trash `R`. Not too bad, right?

.pull-left[
```{r trash-cologne-plot, eval = FALSE}
trash_cologne <- 
  sags_uns_cologne %>% 
  dplyr::filter(field_category_icon == "trash")

tm_shape(trash_cologne) +
  tm_dots()
```
]

--

.pull-right[
```{r trash-cologne-plot-exec, ref.label = "trash-cologne-plot", echo = FALSE}
```
]

---

## Raster Data Access

APIs are not only used for vector data but also raster data.

The idea is the same
- accessing the information through URLs
- just the downloaded data formats differ

---

## OSM Data Can Be Gathered As Raster Data, Too

The `tmaptools` package is pretty handy to download OpenStreetMap data (tiles).

```{r cologne-raster}
cologne_raster <-
  tmaptools::read_osm(trash_cologne) %>% # it's a stars object
  terra::rast()

cologne_raster
```

---

## Mapped OSM Raster Data

The resulting data can be packed in a `tmap` workflow.

.pull-left[
```{r cologne-raster-map, eval = FALSE}
tm_shape(cologne_raster) +
  tm_rgb()
```
]

--

.pull-right[
```{r cologne-raster-map-exec, ref.label = "cologne-raster-map", echo = FALSE}
```
]

---

## Use It As a Background Map

As these data are images, they are perfect to be used as background maps when mapping other geospatial attributes, such as our trash data.

.pull-left[
```{r cologne-raster-map-bg, eval = FALSE}
tm_shape(cologne_raster) +
  tm_rgb() +
  tm_shape(trash_cologne) +
  tm_dots(size = .3)
```
]

--

.pull-right[
```{r cologne-raster-map-bg-exec, ref.label = "cologne-raster-map-bg", echo = FALSE}
```
]

---

## Playing With Different Map Types

List of available type names can be seen with the function call `OpenStreetMap::getMapInfo()`.

--

.pull-left[
```{r esri-topo, out.width = "70%"}
tmaptools::read_osm(
  trash_cologne, 
  type = "bing"
) %>% 
  tm_shape() +
  tm_rgb() +
  tm_shape(trash_cologne) +
  tm_dots(size = .3)
```
]

--

.pull-right[
```{r stamen-watercolor, out.width = "70%"}
tmaptools::read_osm(
  trash_cologne, 
  type = "stamen-watercolor"
) %>% 
  tm_shape() +
  tm_rgb() +
  tm_shape(trash_cologne) +
  tm_dots(size = .3)
```
]

---

## OGC Web Services

OSM and others provide their standardized interface to their data.

The Open Geospatial Consortium developed a more broadly used interface design for more diverse data sources
- often used by public authorities all over the world
- support of well-supported data formats
- good documentation

---

## Displaying vs. Downloading

These web services can broadly be divided into services to
- display data
- download data

Let's briefly focus on download services!

---

## Download Services

Web Feature Service (WFS)
- vector data

**Web Coverage Service (WCS)**
- raster data

Unfortunately, as of today, no ready-to-play packages are providing full access to OWS services in `R`
- The [`ows4R` package](https://cran.r-project.org/web/packages/ows4R/index.html) lets you only use WFS services
- But you could establish an interface to `Python` and use its `OWSLib` to access WCS data (see add-on slides.

---

## Shameless Advertising: The `z11` Package

There are times where no APIs are available for downloading geospatial data.

The German Census 2011 is a prime example
- only a data dump of Gigabytes of data exists for hundreds of attributes
- when you often work with these data, it's a pain

**Thus, I created my own (pseudo-)API for these data hosted over Github.**

---

## Accessing Data

Details on using the package can be found [here](https://stefanjuenger.github.io/z11/articles/using-z11.html), but it's straightforward to use. For example, if you want to download data on immigrant rates on a 1 km² grid, you can use the following function.

```{r census-immigrants}
immigrants_germany <-
  z11::z11_get_1km_attribute(Auslaender_A)

immigrants_germany[immigrants_germany <= -1] <- NA

immigrants_germany
```

---

## It’s a Raster

As it is a raster file, you can plot it easily.

--

.pull-left[
```{r census-immigrants-map, eval = FALSE}
tm_shape(immigrants_germany) +
  tm_raster(palette = "viridis")
```
]

--

.pull-right[
```{r census-immigrants-map-execution, ref.label = "census-immigrants-map", echo = FALSE}
```
]

---

class: middle 
## Exercise 2_1_2: Wrangling the German Census

[Exercise](https://stefanjuenger.github.io/gis_socsci_konstanz/exercises/2_1_2_Wrangling_the_German_Census.html)

[Solution](https://stefanjuenger.github.io/gis_socsci_konstanz/solutions/2_1_2_Wrangling_the_German_Census.html)


<!-- STOP WITH SLIDES HERE -->

---

```{r child = "./content/config/sessions_end.Rmd"}
```