---
title: "Introduction to Geospatial Techniques for Social Scientists in R"
subtitle: "Applied Data Wrangling"
author: "Stefan Jünger & Anne-Kathrin Stroppe"
date: June 07, 2022
institute: |
  GESIS Workshop
presenter: Anne
---
layout: true

```{r child = "./content/config/sessions_setup.Rmd"}
```

---

## Now

```{r course-content-now, echo = FALSE}
course_content %>%
  kableExtra::row_spec(7, background = "yellow")
```

---

## What Are Georeferenced Data?

.pull-left[
</br>
Data with a direct spatial reference $\rightarrow$ **geo-coordinates**
- Information about geometries
- Optional: Content in relation to the geometries
]

.pull-right[
```{r pic-geometries-1, echo = FALSE, out.width="85%"}
woRkshoptools::include_picture("fig_geometries.png")
```

.tinyisher[Sources: OpenStreetMap / GEOFABRIK (2018), City of Cologne (2014), and the Statistical Offices of the Federation and the Länder (2016) / Jünger, 2019]
]

---


## Georeferenced Survey Data

Survey data enriched with geo-coordinates (or other direct spatial references)

</br>

```{r pic-surveys, echo = FALSE, out.width="85%"}
woRkshoptools::include_picture("geo_surveys.png")
```

</br>

.center[**With georeferenced survey data, we can analyze interactions between individual behaviors and attitudes and the environment.**]

---

## An Example Workflow

.pull-left[
From the addresses to analyses with georeferenced survey data, several steps and challenges along the way. We will talk about:

- Data Protection & Data Access
- Geocoding
- Spatial Data Linking
- Applied Examples
]

.pull-right[
```{r pic-varreport, echo = FALSE, out.width="75%"}
woRkshoptools::include_picture("varreport.png")
```
]

---

## Data Protection

</br>
</br>
That‘s one of the biggest issues
- Explicit spatial references increase the risk of re-identifying anonymized survey respondents
- Can occur during the processing of data but also during the analysis

</br>

.center[**Affects all phases of research and data management!**]

## Data Availability

.pull-left[
Geospatial Data
- Often de-centralized distributed 
- Fragmented data landscape, at least in Germany

Georeferenced Survey Data
- Primarily, survey data
- Depends on documentation
- Access difficult due to data protection restrictions
]

.pull-right[
```{r pic-data-availability, echo = FALSE, out.width = "75%"}
woRkshoptools::include_picture("data_availability.png")
```
.right[.tinyisher[
https://www.eea.europa.eu/data-and-maps
https://datasearch.gesis.org/
https://datasetsearch.research.google.com/
]]
]

---

## Distribution & Re-Identification Risk

Data may still be sensitive
- Geospatial attributes add new information to existing data
- May be part of general data privacy checks, but we may not distribute these data as is

.pull-left[
Safe Rooms / Secure Data Centers
- Control access
- Checks output
]

.pull-right[
```{r pic-safe-room, echo = FALSE}
woRkshoptools::include_picture("safe_room.png")
```
.right[.tinyisher[https://www.gesis.org/en/services/processing-and-analyzing-data/guest-research-stays/secure-data-center-sdc]]
]

---

## Legal Regulations in Data Processing

.pull-left[
Storing personal information such as addresses in the same place as actual survey attributes is not allowed in Germany
- Projects keep them in separate locations
- Can only be matched with a correspondence table
- Necessary to conduct data linking
]

.pull-right[
```{r pic-linking-2, echo = FALSE}
woRkshoptools::include_picture("fig_linking_workflow_simple.png")
```

.right[.tinyisher[Jünger, 2019]]
]

---

## Geocoding

Indirect spatial references have to be converted into direct spatial references $\rightarrow$ Addresses to geo-coordinates

Different service providers can be used, e.g., Google, Bing, OSM but raises questions of data protection and quality concerns.

We rely on a service offered by the Federal Agency of Cartography and Geodesy (BKG):

- Online interface and API for online geocoding
- Offline geocoding possible based on raw data
- But: Data and service are restricted

---

## `bkggeocoder`

.pull-left[

R package `bkggeocoder` developed at GESIS for (offline) geocoding by Stefan and Jonas Lieth:

- Access via [Github](https://github.com/StefanJuenger/bkggeocoder)
- Introduction in the [Meet the Experts Talk](https://www.youtube.com/watch?v=ZnA21LyKK88&feature=youtu.be) by Stefan

]

.pull-right[
</br>
</br>
```{r pic-bkggeocoding, out.width = "65%", echo = FALSE}
woRkshoptools::include_picture("bkggeocoder.png")
```
]

---


## Spatial Linking

.pull-left[
Geocoding tool retrieves automatically point coordinates, administrative unit keys and INSPIRE* grid cell ids.

Spatial joins based on coordinates for other units:

- constituencies
- administrative units across time (e.g., harmonized territorial status)

]

.pull-right[

```{r pic-spatial-linking, echo = FALSE, out.width = "80%"}
woRkshoptools::include_picture("fig_3d_.png") 
```

.tinyisher[Sources:  OpenStreetMap / GEOFABRIK (2018), City of Cologne (2014), Leibniz Institute of Ecological Urban and Regional Development (2018), Statistical Offices of the Federation and the Länder (2016), and German Environmental Agency / EIONET Central Data Repository (2016) / Jünger, 2019]
]
---

## Data Linking

Linking via ids most commonly used but comes with its own challenges (e.g., territorial status and land reforms? comparable units? heterogeneity within units?).

```{r pic-data-linking, echo = FALSE, out.width = "75%"}
woRkshoptools::include_picture("data_linking.png") 
```

---

## Spatial Linking Methods (Examples) I

.pull-left[

1:1
.tinyisher[sf::st_join]

```{r pic-noise-by-location, echo = FALSE, out.width = "75%"}
woRkshoptools::include_picture("fig_linking_by_location_noise.png")
```
]

.pull-right[

Distances
.tinyisher[sf::st_distance]

```{r pic-noise-distance, echo = FALSE, out.width = "75%"}
woRkshoptools::include_picture("fig_linking_distance_noise_appI.png")
```
]

.tinyisher[Sources: German Environmental Agency / EIONET Central Data Repository (2016) and OpenStreetMap / GEOFABRIK (2018) / Jünger, 2019]

---

## Spatial Linking Methods (Examples) II

.pull-left[

Filter methods
.tinyisher[sf::st_filter or terra::vect(. , filter = )]

```{r pic-immigrants-filter, echo = FALSE, out.width = "75%"}
woRkshoptools::include_picture("fig_linking_focal_immigrants.png")
```

]

.pull-right[

Buffer zones
.tinyisher[sf::st_buffer (combined with terra::vect())]

```{r pic-selaing-buffer, echo = FALSE, out.width = "75%"}
woRkshoptools::include_picture("fig_linking_buffer_sealing.png")
```
]

.tinyisher[Sources: Leibniz Institute of Ecological Urban and Regional Development (2018) and Statistical Offices of the Federation and the Länder (2016) / Jünger, 2019]

---

## Fake Research Question

.pull-left[
Say we're interested in the impact of the current pandemic on individual well-being in the geographic context.

We plan to conduct a survey in the state of North-Rhine Westphalia.
]

.pull-right[
</br>
```{r pic-trump-fake, echo = FALSE}
woRkshoptools::include_picture("4iq3kg.jpg")
```
.center[.tinyisher[https://imgflip.com/memegenerator/Trump-Bill-Signing]
]
]

---

## Our Sample Area: NRW's Boundaries

.pull-left[
```{r get-nrw}
nrw <-
  osmdata::getbb(
    "Nordrhein-Westfalen", 
    format_out = "sf_polygon"
  ) %>% 
  .$multipolygon %>% 
  sf::st_transform(3035) 

```
]

--

.pull-right[
```{r nrw-map}
tm_shape(sampling_area) +
  tm_borders() 
```
]

---

## A Fake-Life Application

.pull-left[
Let's sample 1,000 people to interview them about their lives.

We can draw a fake sample this way and also add an identifier for the respondents:

```{r set-seed}
set.seed(1234)
```


```{r simulate-coordinates}
fake_coordinates <-
  sf::st_sample(sampling_area, 1000) %>% 
  sf::st_sf() %>% 
  dplyr::mutate(
    id_2 = 
      stringi::stri_rand_strings(10000, 10) %>% 
      sample(1000, replace = FALSE)
  )
```
]

--

.pull-right[
```{r map-osm-coordinates}
tm_shape(sampling_area) +
  tm_borders() +
  tm_shape(fake_coordinates) +
  tm_dots()
```
]

---

## Correspondence Table

As in any survey that deals with addresses, we need a correspondence table of the distinct identifiers.

```{r create-correspondence-table}
correspondence_table <-
  dplyr::bind_cols(
    id = 
      stringi::stri_rand_strings(10000, 10) %>% 
      sample(1000, replace = FALSE),
    id_2 = fake_coordinates$id_2
  )

correspondence_table
```

---

## Conduct the Survey

We ask respondents for some standard sociodemographics. But we also apply a new and highly innovative item score, called the Fake Corona Burden Score (FCBS) using the [`faux` package](https://cran.r-project.org/web/packages/faux/index.html).

```{r hidden-data, include = FALSE}
secret_data <- readRDS("./data/secrect_data.rds")

secret_variable_we_are_hiding_from_you <-
  faux::rnorm_pre(
        secret_data$pop_dens, 
        mu = 50, 
        sd = 10, 
        r = -0.5
      )
```

```{r simulate-survey-data}
fake_survey_data <- 
  dplyr::bind_cols(
    id = correspondence_table$id,
    age = sample(18:100, 1000, replace = TRUE),
    gender = 
      sample(1:2, 1000, replace = TRUE) %>% 
      as.factor(),
    education =
      sample(1:4, 1000, replace = TRUE) %>% 
      as.factor(),
    income =
      sample(100:10000, 1000, replace = TRUE),
    fcbs = secret_variable_we_are_hiding_from_you
  )
```

---

## Survey Data Structure

```{r show-survey-data}
fake_survey_data 
```

---

## What could explain our Fake Corona Burden Score?

*Likelihood to meet people*
> Higher district's population density, lower Fake Corona Burden Score.

--

*Provision of health services*
> Higher distance to closest hospital,  higher  Fake Corona Burden Score.

--

*Possible language issues in health care communication*
> Higher immigrant rate in the neighborhood, higher Fake Corona Burden Score.

---

## Population Density

When all data sets are loaded, we reduced our sample to the area North Rhine Westphalia.

```{r load-data-hidden, echo = F, warning = F, message = F}

sampling_area_districts <-
  sf::read_sf("./data/VG250_KRS.shp") %>% 
  sf::st_transform(3035) %>% 
  # reduce to area of nrw: x lies within y
  sf::st_join(.,
              sampling_area, 
              join = sf::st_intersects, 
              left = FALSE) %>% 
  # select and rename columns
  dplyr::rename(district_id = AGS) %>% 
  dplyr::select(district_id)

attributes_districts <- readr::read_csv("./data/attributes_districts.csv")
 
sampling_area_districts_enhanced <-
  sampling_area_districts %>% 
  dplyr::left_join(attributes_districts, by = "district_id" )

hospitals <-
  readr::read_csv("./data/hospital_points.csv") %>%
  sf::st_as_sf(coords = c("X", "Y"), crs = 3035)

sampling_area_hospitals <-
  hospitals %>% 
  sf::st_join(
    sampling_area_districts_enhanced %>% 
      dplyr::select(geometry),
    join = st_within
  )

```


```{r prep_data, eval = F}

# reduce sample
sampling_area_districts_enhanced <-
  # load shapefile
  sf::read_sf("./data/VG250_KRS.shp") %>% 
  # adjust crs
  sf::st_transform(3035) %>% 
  # add attributes
  dplyr::left_join(attributes_districts, by = "district_id" ) %>% 
  # reduce to area of nrw: x lies within y
  sf::st_join(.,
              sampling_area, 
              join = sf::st_intersects, 
              left = FALSE) %>% 
  # select and rename columns
  dplyr::rename(district_id = AGS) %>% 
  dplyr::select(district_id)


```

---

## Calculate size of the area

```{r area-calc}

# calculate area of districts
# areas will always be calculated
# in units according to the CRS 
sf::st_area(sampling_area_districts_enhanced) %>% 
  head(4)

```


---

## Population Density
All left to do is a simple mutation. Let's pipe it!

.pull-left[
```{r pop-dens-pipe, eval = F}
# calculation population density
sampling_area_districts_enhanced <-
  sampling_area_districts_enhanced %>% 
  # calculate area of districts (areas will always
  # be calculatednin units according to the CRS )
  dplyr::mutate(area = sf::st_area(.)) %>% 
  # change unit to square kilometers
  dplyr::mutate(area_km2 = units::set_units
                    (area, km^2)) %>% 
  # recode variable as numeric
  dplyr::mutate(area_km2 = as.numeric
                    (area_km2)) %>% 
  # calculate population density
  dplyr::mutate(pop_dens = population/
           area_km2)
```
]

.pull-right[
```{r, echo = F}
sampling_area_districts_enhanced <-
  sampling_area_districts_enhanced %>% 
  # calculate area 
  dplyr::mutate(area = sf::st_area(.)) %>% 
  # change unit to square kilometers
  dplyr::mutate(area_km2 = units::set_units
                    (area, km^2)) %>% 
  # recode variable as numeric
  dplyr::mutate(area_km2 = as.numeric
                    (area_km2)) %>% 
  # calculate population density
  dplyr::mutate(pop_dens = population/
           area_km2)

tm_shape(sampling_area_districts_enhanced) +
  tm_fill("pop_dens", breaks = c(0,100,200,400,800,1600,3200, Inf))+
  tm_layout(legend.outside = TRUE)
```
]

---

## Respondents in Districts

We have population density on the district level. Since our analysis focuses on the individual-level, we can spatial join the information to our fake respondents' coordinates.

```{r join-disp}

# join back
spatial_information <-
  sampling_area_districts_enhanced %>%
  # keeping just the variables we want
  dplyr::select(district_id, pop_dens) %>% 
  # since we want to join district to
  # respondent defining coordintes first
  sf::st_join(fake_coordinates,
  # district data second
          . ,
    # some points may lie on the border
    # choosing intersects therefore
          join = sf::st_intersects) %>% 
  # drop our coordinates for data protection
  sf::st_drop_geometry()

```

---

## Respondents in Districts

```{r join-disp-print}
head(spatial_information, 5)
```

---

# Distance Calculation

`sf::st_distance()` will calculate between **all** respondents and **all** hospitals resulting in a matrix with 1,786,000 objects (1,000 respondent * 1,786 hospitals). We can make our lives a little bit easier by treating this matrix as a `tibble`.

.pull-left[
```{r dist_matr-disp, eval = F}
# distances between each respondent 
# and each hospital
distance_matrix <- 
  # point layer "distance from" 
  sf::st_distance(
    fake_coordinates, 
    # point layer "distance to"
    sampling_area_hospitals,
    # dense matrix with all 
    # pairwise distance
    by_element = FALSE
  ) %>% 
  # making life a little bit easier
  dplyr::as_tibble() 

# check our matrix
# again, units = CRS units!
distance_matrix
```
]

.pull-right[
```{r dist-matr, echo = F}
# calculate distances between each respondent and each hospital
# point layer "distance from" 
sf::st_distance(
  fake_coordinates, 
  # point layer "distance to"
  sampling_area_hospitals,
  # returns dense matrix with all pairwise distance
  by_element = FALSE) %>% 
  dplyr::as_tibble()
```
]

---

## Find Minimum Distance

That's all there is concerning the "spatial" part of our data wrangling. 
From now on, just good old (boring) data crunching to get our distance to the closest hospital.

.pull-left[
```{r closest-hosp-disp, eval = F}
distance_closest <-
    distance_matrix %>% 
    # from unit to numeric
    dplyr::mutate_all(as.numeric) %>% 
    # identify for each row the minimum
    # & save in variable
    dplyr::mutate(dist_closest_hospital = 
                  (apply(., 1, min))) %>%  
    # get kilometer instead of meter
    dplyr::mutate(dist_closest_hospital = dist_closest_hospital/1000 ) %>%
    # select only column 
    # containing smallest distance
    dplyr::select(dist_closest_hospital)
```
]

.pull-right[
```{r closest-hosp, echo = F}
sf::st_distance(
  fake_coordinates, 
  sampling_area_hospitals,
  by_element = FALSE
) %>% 
  dplyr::as_tibble() %>% 
  dplyr::mutate_all(as.numeric) %>% 
  # identify for each row the minimum & save var
  dplyr::mutate(dist_closest_hospital = (apply(., 1, min))) %>% 
  dplyr::mutate(dist_closest_hospital = dist_closest_hospital/1000) %>% 
  # select only column containing smallest distance
  dplyr::select(dist_closest_hospital)

```

]

---

## Join to our spatial information!

I prefer to work with kilometers rather than meters. And I want to add our new variable to the other spatial information we already prepared. Luckily, I know that the spatial information table has the same length and order as the fake coordinates.

.pull-left[
```{r dist-resp-disp, eval = F}
spatial_information <-
  distance_closest %>% 
  # bind columns with spatial information
  # only bind with other data set than the
  # original coordinates when you are 100
  # percent sure it's same length and order!
  dplyr::bind_cols(spatial_information, .) 
```
]

.pull-right[
```{r dist-resp, echo = F}
spatial_information <-
  sf::st_distance(
    fake_coordinates, 
    sampling_area_hospitals,
    by_element = FALSE
  ) %>% 
  dplyr::as_tibble() %>% 
  dplyr::mutate_all(as.numeric) %>% 
  dplyr::mutate(dist_closest_hospital = (apply(., 1, min))) %>%  
  dplyr::select(dist_closest_hospital) %>% 
  dplyr::bind_cols(spatial_information, .) 

head(spatial_information,6)
```
]

---

## Immigrant Rate Buffers

...and we're not yet done: we still need the immigrant rate in the neighborhood. Let's calculate buffers of 500 meters and add their mean values to our dataset.


```{r load-link-data, include = FALSE}
# download data & create rate

immigrants <- 
  z11::z11_get_100m_attribute(STAATSANGE_KURZ_2)
inhabitants <-
  z11::z11_get_100m_attribute(Einwohner) %>% 
  terra::resample(immigrants)
immigrant_rate <- immigrants / inhabitants * 100
# calculate immigrant rate for 500m buffer
immigrant_buffers <- 
terra::extract(
  immigrant_rate, 
  fake_coordinates %>% 
    sf::st_buffer(500) %>% 
    terra::vect(), 
  fun = mean,
  na.rm = TRUE
  )
# spatially link with buffers on the fly
spatial_information <-
  spatial_information %>% 
  dplyr::mutate(immigrant_buffers = immigrant_buffers[[2]])

```

```{r load-link-noeval, eval = FALSE}

# download data & create rate
immigrants <- 
  z11::z11_get_100m_attribute(STAATSANGE_KURZ_2)
inhabitants <-
  z11::z11_get_100m_attribute(Einwohner) %>% 
  terra::resample(immigrants)
immigrant_rate <- immigrants / inhabitants * 100
# calculate immigrant rate for 500m buffer
immigrant_buffers <- 
terra::extract(
  immigrant_rate, 
  fake_coordinates %>% 
    sf::st_buffer(500) %>% 
    terra::vect(), 
  fun = mean,
  na.rm = TRUE
  )
# spatially link with buffers on the fly
spatial_information <-
  spatial_information %>% 
  dplyr::mutate(immigrant_buffers = immigrant_buffers[[2]])

```

```{r save-secret-data, echo = FALSE, include = FALSE}
saveRDS(spatial_information, "./data/secrect_data.rds")
```


---

## Join with Fake Burden Score

I hope you're not tired to join data tables.
Since we care a tiny bit more about data protection than others, we have yet another joining task left: joining the information we received using our (protected) fake coordinates to the actual survey data via the correspondence table.

.pull-left[
```{r spatial-info-join}
# last joins for now
fake_survey_data_spatial <-
  # first join the id
  dplyr::left_join(
    correspondence_table, 
    spatial_information, 
    by = "id_2"
  ) %>% 
  # drop the fake_coordinate id
  dplyr::select(-id_2) %>% 
  # join the survey data
  dplyr::left_join(
    fake_survey_data,
    by = "id"
  ) 
```
]

.pull-right[
```{r correlation-plot, echo = FALSE, out.width = "75%"}
fake_survey_data_spatial %>% 
  dplyr::select(
    fcbs, pop_dens, dist_closest_hospital, immigrant_buffers
  ) %>% 
  corrr::correlate() %>% 
  corrr::network_plot(min_cor = .1)
```

]

---

class: middle
## Exercise 2_2_1 (optional!): Spatial Joins

[Exercise](https://stefanjuenger.github.io/gesis-workshop-geospatial-techniques-R-2023/exercises/2_2_1_Spatial_Joins_OPTIONAL.html)

[Solution](https://stefanjuenger.github.io/gesis-workshop-geospatial-techniques-R-2023/exercises/2_2_1_Spatial_Joins_OPTIONAL.html)


---

class: middle

## Addon-slides: Example Studies

---

## Environmental inequalities (Jünger, 2021)

> Is income associated with fewer environmental disadvantages, and are there differences between German people and people with a migration background?

.pull-left[
.small[
Theoretical Framework
- Social and Ethnic Inequalities (Crowder & Downey, 2010)
- Place Stratification (Lersch, 2013)

Data
- GGSS 2016 & 2018
- soil sealing & green spaces
]
]

.pull-right[
```{r pic-linking-appIII, echo = FALSE, out.width = "65%"}
woRkshoptools::include_picture("fig_linking_buffer_sealing.png")
```

.tinyisher[Leibniz Institute of Ecological Urban and Regional Development (2018) / Jünger, 2019]
]


---

## Results

```{r pic-soil-sealing-estimates, echo = FALSE, out.width = "70%"}
woRkshoptools::include_picture("FIGURE_2.png")
```

.tinyisher[Data source: GGSS 2016 & 2018; N = 6,117; 95% confidence intervals based on cluster-robust standard errors (sample point); all models control for age, gender, education, household size, german region and survey year interaction, inhabitant size of the municipality, and distance to municipality administration]

---

## Attitudes towards minorities (Jünger & Schaeffer, 2022)

> Do people who live in ethnic homogenous neighborhoods that are close to ethnic diverse ones have more negative attitudes towards minorities?

.pull-left[
.small[
Theoretical Framework
- Contact Theory (Allport, 1954)
- Ethnic Competition (Stephan et al., 2009)

Data
- GGSS 2016
- German Census 2011
]
]

.pull-right[
```{r halo-ii, echo = FALSE, out.width = "65%"}
woRkshoptools::include_picture("Abb1.png")
```

.tinyisher[German Census 2011, OpenStreetMap / Jünger & Schaeffer, 2022]
]

---

## Results

```{r halo-ii-2, echo = FALSE, out.width = "70%"}
woRkshoptools::include_picture("Abb2.png")
```

.tinyisher[Data source: GGSS 2016; N = 1,689; 95% confidence intervals based on cluster-robust standard errors (sample point); all models control for age, gender, education, income, unemployment, homeownership, immigrants and inhabitants in the neighborhood, inhabitant size of the municipality, german region]

---

## Left Behind by the State? (Stroppe, Forthcoming)

> Are political trust levels affected by the accessibility of public services and infrastructures for citizens?

.pull-left[
.small[
Theoretical Framework
- Political Performance-Trust Link (Easton 1965, Hetherington 2005)
- Context condition low-intensity information cue (Cho & Rudolph 2008)

Data
- GGSS 2018
- hospital, school, train station (distance measures)
- municipality data
]
]

.pull-right[
<br>
```{r pic-provision, echo = FALSE, out.width = "65%"}
woRkshoptools::include_picture("meandist_trains.png")
```

.tinyisher[Federal Statistical Office 2019, Deutsche Bahn 2017 and GeoBasis-DE / BKG 2022 / Stroppe, 2023]
]


---

## Results

<br>
```{r pic-coefplotfig1, echo = FALSE, out.width = "95%"}
woRkshoptools::include_picture("fig1_coefplot_colored.png")
```

.tinyisher[Data source: GGSS 2018 and Federal Statistical Office 2017. N = 3030,  Groups = 152 (Municipalities). Fitted Models: OLS multi-level random effect models. Individual-level controls: income, gender, education, age, personal trust, political interest. Municipality level controls: population density and unemployment. Dependent variable: Trust in government. Survey weights are applied.]

---

```{r child = "./content/config/sessions_end.Rmd"}
```